Blame and Coercion: Together Again for the Second Time

### 2019-10-16 paper.tex
edits

@@

\citet{siek2009exploring} decorate \citet{herman2010space}'s coercions with 
labels to restore 
blame-tracking. They also propose that there are four blame 
strategies for their coercion:
$ \{Lazy, Eager\} \times \{D, UD\} $. 
Eager strategies detect more potential type errors, 
but blame more programs.
$ D $ and $ UD $ assign blame labels differently.
They prove partially that their \lazyUD{} coercion is correct by showing 
that it simulates the Blame Calculus in \cite{wadler2009well}. It is unknown, 
however, whether the simulation also works in the other direction.

@@

\citet{siek2012interpretations} present the first \lazyD{} Blame Calculus and 
revised coercion calculi.
They simplify the coercion calculi by only 
working with normal coercions (i.e. coercions in normal form). 
They conjecture that their revised coercion calculi bisimulate the 
corresponding Blame Calculi.

Following \citet{siek2010threesomes}, \citet{garcia2013calculating} develops
threesome for other blame strategies. After studying in depth the relation 
between coercion and threesome,
he claims that normal coercion with labels is easy to understand but hard to 
implement and hard to reason about, and that 
threesome with labels, however, is easy to implement and easy to reason about 
but hard to understand. 
His claim is mostly affirmed by the group of people who develop threesome 
(\citet{siek2015blame}).
His threesome calculi are equivalent to the coercion calculi in 
\citet{siek2009exploring} because the former ones are derived from the latter 
ones.

\citet{siek2015blame} revisit the coercion-based approach. They revise the 
\lazyUD{} coercion in \citet{siek2012interpretations} and prove that 
their revised coercion calculus bisimulates the \lazyUD{} Blame Calculus.

@@

So far, it seems that normal coercion is the best way to combine 
space-efficiency and blame-tracking: normal coercion is claimed easy to 
understand \cite{garcia2013calculating}\cite{siek2015blame} and is shown 
easy enough to implement in a compiler \cite{kuhlenschmidt2018efficient}.
Normal coercion is still not satisfactory, however, in at least two aspects.
Firstly, as we mentioned above, it is hard to reason about normal coercion. The 
major difficulty is from its non-structurally recursive composition.
For instance, many developers of Grift report it is tricky to convince
their proof assistants that the composition terminates. 
Secondly, the definition of normal coercion lies unnecessarily on top 
of coercion: all coercions are normal when they are initially constructed, and 
\citet{siek2012interpretations} have shown an function that 
composes and produces normal coercions.

Perhaps unsurprisingly, threesome has a structurally recursive compose and a 
self-standing definition. Together with \citet{garcia2013calculating}'s 
work, they suggest that there might be a cast representation whose definition 
is self-standing, and whose composition is structurally recursive. 
Super-coercion introduced in \citet{garcia2013calculating} could be a promising 
candidate. It is believed even more complicated than threesome 
\cite{siek2015blame}. In fact, it uses 10 constructors to deal with an 
elementary type system with only base types and 
function types. And four constructors are directly related to function types. 
Thus, super-coercion might not scale very well to more sophisticated type 
systems.

We present hypercoercion, a cast representation whose composition is 
structurally recursive, and whose definition is self-standing. What's more, it 
should be at 
least as easy to implement and understand as normal coercions, because there is 
a clear connection between them, as we will show in Section 
\ref{sec:hypercoercion}.
Our hypercoercion considers sum types and product types, which are not 
considered in all proofs above. Adding each of them requires us to add only one 
new constructor to a component of hypercoercion. This suggests that 
hypercoercion might scale better than super-coercion.

We prove formally that the \lazyD{} (resp. \lazyUD{}) hypercoercion calculus 
bisimulates the \lazyD{} (resp. \lazyUD{}) Blame Calculus. This is 
the first bi-simulation proof for the \lazyD{} Blame Calculus as far as we know.

The structure of this paper is as follows. 
Section \ref{sec:blame-calculus} reviews the state-of-art of $Lazy$ Blame 
Calculi. 
Section \ref{sec:coercion-calculus} reviews the state-of-art of $Lazy$ Coercion 
Calculi. 
In section \ref{sec:hypercoercion} we present hypercoercion.
Section \ref{sec:conclude} concludes.

\section{Background: Blame Calculus} \label{sec:blame-calculus}

@@

	\caption{Blame Calculus and its static semantics}
	\label{fig:blame-static}
\end{figure}

@@

%One might be tempted to conclude that inconsistency is the root of all runtime 
%type errors (blames). In the setting of gradual typing, however, runtime 
%type-checking is usually gradual as well. 
$ S \smile T $ reads ``$ S $ and $ T $ are shallowly-consistent''. Two types 
are shallowly consistent if one of them is $ \star $, or they have the same 
top-most type constructor. Shallow-inconsistency is the root of all blames in 
all lazy strategies -- casting a value to a shallowly inconsistent type leads 
to a blame. Shallow-consistency is reflexive, symmetric, but 
not transitive.

Let $ e $ ranges over terms, including all terms from $ \mathtt{STLC+} $ and, 
in addition, casts and blames. Unlike $ \mathtt{STLC+} $, we annotate the 
co-domain of lambda abstractions explicitly. 
The end of the figure shows the typing rules for the additional terms.

Let $ o $ ranges over observations. They are what would be observed if a 
program terminates. Observations include all constructors and blames.

@@

\subsection{\lazyD{} Blame Calculus}

@@

Fig.~\ref{machine-cekc} defines the dynamic semantics of the \lazyD{} Blame 
Calculus in the style of CEK machines (\citet{felleisen1986control}).
\todo[inline]{The figure is too tall. You may need to pull some things
  out of the figure, such as the value typing. -Jeremy}
\todo[inline]{Please fill in the missing reduction rules. -Jeremy}
\todo[inline]{Explain why you are using a CEK machine instead of
  reduction semantics. -Jeremy}
\todo[inline]{Change the order of parameters for applyCast to take
  the value first and the cast second. -Jeremy}
\todo[inline]{The definition of values is too loose for a cast.
  You need to restrict the casts to just the inert ones. -Jeremy }
We conjecture that this CEK machines agree with the \lazyD{} blame calculus in 
\citet{siek2012interpretations}.
We will show in \hyperref{sec:hypercoercion} that this abstract machine is 
equivalent to an abstract machine that represents casts with hypercoercions.

Let $ c $ ranges over casts. A cast is a triple of a type, a label, and a type.
Let $ v $ ranges over values. A value is either a function, the 
element of unit, a pair, a left injection, a right injection, or a casted value.
The cast around a casted value is not arbitrary, as shown by value typing: the 
source type must be a pre-type, and the source and the target must be 
shallow-consistent.
Let $ r $ ranges over cast results. A cast result is either a success, which 
brings a value, or a failure, which brings a blame label.
Let $ s $ ranges over machine states. A state is either looking at an 
expression to decide what to do next, returning a value to a continuation, or 
halting with an observation.
Let $ \kappa $ ranges over continuations. $ \mathtt{mt} $ is the top 
continuation. The remaining continuations correspond to expressions. For 
example, $ (\mathtt{cons_1} \; e \; \rho \; \kappa) $ is the continuation where 
we are waiting for the value of the first argument to a $ \mathtt{cons} $. And 
the last continuation, $ \langle c \rangle \kappa $ is to cast the value before 
returning to $ \kappa $.

In the reduction rules, to evaluate a cast expression, the machine move the 
cast to the continuation and evaluate the inner expression. To apply a casted 
function, the machine firstly cast $ v_1 $, the operand, then apply the casted 
operand to $ v_2 $, the un-casted function, and finally cast the result of the 
function application. To take out the first (resp. second) part of a casted 
pair, the machine firstly take out the first (resp. second) part of $ v $, the 
un-casted pair, and cast the result. To case-split a sum, the machine looks 
deep inside the value to find out whether it is a left injection or a right 
one. Along the way, the machine accumulate the pending casts onto the 
continuations. To cast a value, the machine try to apply the cast to the value. 
If the cast succeeds, the machine returns the result to the next continuation. 
If the cast fails, the machine halts with the blame label.

@@

\subsection{\lazyUD{} Blame Calculus [Jeremy]}

UNDER CONSTRUCTION

@@

\begin{proposition}[Strong Bi-simulation among $ \mathcal{S}(\cdot) $]
	\label{thm:CEKS-bisim}
	If 
	$ C_1 $ and $ C_2 $ are surely \lazyD,
	$ s_1, s_2 \in S(C_1) $ and $ s_3 \in S(C_2) $,
	$ s_1 \approx s_3 $,
	$ \judgeSreduce{C_1}{s_1}{s_2} $,
	then there exists an $ s_4 $ such that
	\begin{itemize}
	\item $ s_2 \approx s_4 $ and
	\item $ s_3 \longrightarrow_{S(C_2)} s_4 $
	\end{itemize}
\end{proposition}
\begin{proof}
	The key ideas of this proof are undoing sequencing with the property (2) of 
	surely \lazyD{} CR, and handling all possibly uses of $ 
	cast(T,l,T) $ with property (3)-(11).
\end{proof}

@@

\begin{lemma}[Weak Bi-simulation between $ \mathcal{S}(\cdot) $ and $ 
\mathcal{D} $]
\label{thm:surely-monoidic-reduce}
For all cast representations $ C $, 
if $ C $ is surely \lazyD{} and monoidic,
$ s_1 \in \mathcal{D} $,
$ s_2 \in \mathcal{S}(C) $, and
$ s_1 \approx s_2 $, 
\begin{enumerate}
	\item If $ \judgeDBreduce{s_1}{s_3} $,
	then
	there exist $ s_5 \in \mathcal{D} $, $ s_4 \in \mathcal{S}(C) $,
	such that \begin{itemize}
		\item $ s_5 \approx s_4 $, and
		\item $ \judgeSreduce{C}{s_2}{s_4} $,
		\item $ \judgeDBreduceTrans{s_3}{s_5} $
	\end{itemize}
	\item If $ \judgeSreduce{C}{s_2}{s_4} $,
	then
	there exist $ s_3 \in \mathcal{D} $,
	such that \begin{itemize}
		\item $ s_3 \approx s_4 $
		\item $ \judgeDBreduceTrans{s_1}{s_3} $
	\end{itemize}
\end{enumerate}
\end{lemma}
\begin{proof}
	See the supplementary material.
\end{proof}

>>>>

\todo[inline]{I see lots of mistakes regarding when to make words
  plural (whether they should have an ``s'' at the end). Please review
  the rules of English regarding plurals. The rules are tricky...
  some words don't have a plural form, like ``blame''. -Jeremy}

\todo[inline]{The first paragraph of the introduction should summarize
  the entire introduction, and in some sense, the entire paper. -Jeremy}

\todo[inline]{BTW, using the Coercion and Blame paper as a model for
  how to write papers is perhaps not the best idea. That paper intentionally
  diverged from the normal paper format in several ways. You can
  do that too after you've become a master at writing papers.
  For now you are much better off following the advice of
  Simon Peyton Jones about how to write a paper (google it).}

@@

\citet{siek2009exploring} decorate \citet{herman2010space}'s coercions
with labels to restore blame-tracking. They also propose that there
are four blame strategies for coercions: $ \{Lazy, Eager\} \times \{D,
UD\} $.  Eager strategies detect more potential type errors, but blame
more programs.  $ D $ and $ UD $ assign blame labels differently.
They prove that the \lazyUD{} calculus simulates the Blame Calculus of
\citet{wadler2009well}.  They do not prove the other direction,
whether the Blame Calculus simulates the \lazyUD{} calculus.
\todo[inline]{Is it important to this paper whether there is a
  simulation in the other direction? -Jeremy}

@@

\citet{siek2012interpretations} present the first \lazyD{} Blame Calculus and 
revised coercion calculi.
\todo[inline]{Not ``first''... lazy D is in the 2009 paper, right?
  Or did you mean something else? -Jeremy}
\todo[inline]{The ``Blame Calculus'' is a \lazyUD{} cast calculus.
  There is no such thing as a \lazyD{} Blame Calculus. -Jeremy}
They simplify the coercion calculi by only 
working with normal coercions (i.e. coercions in normal form). 
They conjecture that their revised coercion calculi bisimulate the 
corresponding Blame Calculi.

Following \citet{siek2010threesomes}, \citet{garcia2013calculating} develops
threesome for other blame strategies. After studying in depth the relation 
between coercion and threesome,
he claims that normal coercion with labels is easy to understand but hard to 
implement and hard to reason about, and that 
threesome with labels, however, is easy to implement and easy to reason about 
but hard to understand. 
His claim is mostly affirmed by the group of people who develop threesome 
(\citet{siek2015blame}).
His threesome calculi are equivalent to the coercion calculi of
\citet{siek2009exploring} because the former ones are derived from the latter 
ones.

\citet{siek2015blame} revisit the coercion-based approach. They revise
the \lazyUD{} coercion in \citet{siek2012interpretations} and prove
that their revised coercion calculus bisimulates the \lazyUD{} Blame
Calculus \citet{wadler2009well}.

@@

So far, it seems that normal coercion is the best way to combine 
space-efficiency and blame-tracking: normal coercion is claimed easy to 
understand \cite{garcia2013calculating}\cite{siek2015blame} and is shown 
easy enough to implement in a compiler \cite{kuhlenschmidt2018efficient}.
Normal coercion is still not satisfactory, however, in at least two aspects.
Firstly, as we mentioned above, it is hard to reason about normal coercion. The 
major difficulty is from its non-structurally recursive composition.
For instance, many developers of Grift report it is tricky to convince
their proof assistants that the composition terminates.
%
\todo[inline]{Don't lie. There are not many developers of Grift.
  I only know 2 people developing Grift: Andre and Deyaa. -Jeremy}
%
Secondly, the definition of normal coercion lies unnecessarily on top 
of coercion: all coercions are normal when they are initially constructed, and 
\citet{siek2012interpretations} have shown an function that 
composes and produces normal coercions.
%
\todo[inline]{Many researchers would view this as a positive thing,
  not a negative. For example, the notion of normal forms comes up
  over and over again in mathematics and logic. -Jeremy}

Perhaps unsurprisingly, threesome has a structurally recursive compose and a 
self-standing definition. Together with \citet{garcia2013calculating}'s 
work, they suggest that there might be a cast representation whose definition 
is self-standing, and whose composition is structurally recursive. 
supercoercion introduced in \citet{garcia2013calculating} could be a promising 
candidate. It is believed even more complicated than threesome 
\cite{siek2015blame}. In fact, it uses 10 constructors to deal with an 
elementary type system with only base types and 
function types. And four constructors are directly related to function types. 
Thus, supercoercion might not scale very well to more sophisticated type 
systems.

We present hypercoercion, a cast representation whose composition is 
structurally recursive, and whose definition is self-standing. What's more, it 
should be at 
least as easy to implement and understand as normal coercions, because there is 
a clear connection between them, as we will show in Section 
\ref{sec:hypercoercion}.
Our hypercoercion considers sum types and product types, which are not 
considered in all proofs above. Adding each of them requires us to add only one 
new constructor to a component of hypercoercion. This suggests that 
hypercoercion might scale better than supercoercion.

We prove formally that the \lazyD{} (resp. \lazyUD{}) hypercoercion calculus 
bisimulates the \lazyD{} (resp. \lazyUD{}) Blame Calculus. This is 
the first bisimulation proof for the \lazyD{} Blame Calculus as far as we know.

\todo[inline]{Explain why we care about \lazyD{} and \lazyUD{}. -Jeremy}

The structure of this paper is as follows. 
Section \ref{sec:blame-calculus} reviews the state-of-art of $Lazy$ Blame 
Calculi. 
Section \ref{sec:coercion-calculus} reviews the state-of-art of $Lazy$ Coercion 
Calculi. 
In section \ref{sec:hypercoercion} we present hypercoercion.
Section \ref{sec:conclude} concludes.
\todo[inline]{Lots of missing sections. -Jeremy}

\section{Background: Cast Calculi} \label{sec:blame-calculus}

@@

	\caption{Syntax and static semantics of the cast calculi.}
	\label{fig:blame-static}
\end{figure}

@@

%One might be tempted to conclude that inconsistency is the root of all runtime 
%type errors (blames). In the setting of gradual typing, however, runtime 
%type-checking is usually gradual as well. 
$ S \smile T $ reads ``$ S $ and $ T $ are shallowly-consistent''. Two types 
are shallowly consistent if one of them is $ \star $, or they have the same 
top-most type constructor. Shallow-inconsistency is the root of all blame in 
lazy cast strategies -- casting a value to a shallowly inconsistent type leads 
to a blame. Shallow-consistency is reflexive, symmetric, but 
not transitive.

Let $ e $ ranges over terms, including all terms from $ \mathtt{STLC+} $ and, 
in addition, casts and blame. Unlike $ \mathtt{STLC+} $, we annotate the 
codomain of lambda abstractions explicitly.
\todo[inline]{Why do you annotate the codomain of lambda abstractions
  explicitly? -Jeremy}
The end of \figref{fig:blame-static} shows the typing rules for the
additional terms.

Let $ o $ ranges over observations. They are what would be observed if a 
program terminates. Observations include all constructors and blames.
\todo[inline]{``blames'' is incorrect. See
https://www.wordhippo.com/what-is/the-plural-of/blame.html}

@@

\subsection{The Blame Calculus (a \lazyD{} cast calculus)}

@@

Fig.~\ref{machine-cekc} defines the dynamic semantics of the \lazyD{} Blame 
Calculus in the style of CEK machines \citep{felleisen1986control}.
\todo[inline]{Briefly present the main ideas regarding how a CEK
  machine works. (2-3 sentences) -Jeremy}
\todo[inline]{Use citep to put a citation in parentheses, as I've
  done above. -Jeremy }
\todo[inline]{The figure is too tall. You may need to pull some things
  out of the figure, such as the value typing. -Jeremy}
\todo[inline]{Please fill in the missing reduction rules. -Jeremy}
\todo[inline]{Explain why you are using a CEK machine instead of
  reduction semantics. -Jeremy}
\todo[inline]{Change the order of parameters for applyCast to take
  the value first and the cast second. -Jeremy}
\todo[inline]{The definition of values is too loose for a cast.
  You need to restrict the casts to just the inert ones. -Jeremy }
We conjecture that this CEK machines agree with the \lazyD{} blame calculus in 
\citet{siek2012interpretations}.
We will show in \hyperref{sec:hypercoercion} that this abstract machine is 
equivalent to an abstract machine that represents casts with hypercoercions.

Let $ c $ ranges over casts.
\todo[inline]{``ranges'' should be ``range'', not plural. -Jeremy}
A cast is a triple of a type, a label, and a type.
Let $ v $ ranges over values. A value is either a function, the 
element of unit, a pair, a left injection, a right injection, or a casted value.
The cast around a casted value is not arbitrary, as shown by value typing: the 
source type must be a pre-type, and the source and the target must be 
shallow-consistent.
Let $ r $ ranges over cast results. A cast result is either a success, which 
brings a value, or a failure, which brings a blame label.
Let $ s $ ranges over machine states. A state is either looking at an 
expression to decide what to do next, returning a value to a continuation, or 
halting with an observation.
Let $ \kappa $ ranges over continuations. $ \mathtt{mt} $ is the top 
continuation. The remaining continuations correspond to expressions. For 
example, $ (\mathtt{cons_1} \; e \; \rho \; \kappa) $ is the continuation where 
we are waiting for the value of the first argument to a $ \mathtt{cons} $. And 
the last continuation, $ \langle c \rangle \kappa $ is to cast the value before 
returning to $ \kappa $.

In the reduction rules, to evaluate a cast expression, the machine move the 
cast to the continuation and evaluate the inner expression. To apply a casted 
function, the machine firstly cast $ v_1 $, the operand, then apply the casted 
operand to $ v_2 $, the un-casted function, and finally cast the result of the 
function application. To take out the first (resp. second) part of a casted 
pair, the machine firstly take out the first (resp. second) part of $ v $, the 
un-casted pair, and cast the result. To case-split a sum, the machine looks 
deep inside the value to find out whether it is a left injection or a right 
one.
\todo[inline]{Change the rules for case to be small step, i.e.,
 the case function should not be recursive. -Jeremy}
Along the way, the machine accumulate the pending casts onto the 
continuations. To cast a value, the machine try to apply the cast to the value. 
If the cast succeeds, the machine returns the result to the next continuation. 
If the cast fails, the machine halts with the blame label.

@@

\subsection{The \lazyUD{} cast calculus [Jeremy]}

@@

\begin{proposition}[Strong bisimulation among $ \mathcal{S}(\cdot) $]
	\label{thm:CEKS-bisim}
	If 
	$ C_1 $ and $ C_2 $ are surely \lazyD,
	$ s_1, s_2 \in S(C_1) $ and $ s_3 \in S(C_2) $,
	$ s_1 \approx s_3 $,
	$ \judgeSreduce{C_1}{s_1}{s_2} $,
	then there exists an $ s_4 $ such that
	\begin{itemize}
	\item $ s_2 \approx s_4 $ and
	\item $ s_3 \longrightarrow_{S(C_2)} s_4 $
	\end{itemize}
\end{proposition}
\begin{proof}
	The key ideas of this proof are undoing sequencing with the property (2) of 
	surely \lazyD{} CR, and handling all possibly uses of $ 
	cast(T,l,T) $ with property (3)-(11).
\end{proof}

@@

\begin{lemma}[Weak bisimulation between $ \mathcal{S}(\cdot) $ and $ 
\mathcal{D} $]
\label{thm:surely-monoidic-reduce}
For all cast representations $ C $, 
if $ C $ is surely \lazyD{} and monoidic,
$ s_1 \in \mathcal{D} $,
$ s_2 \in \mathcal{S}(C) $, and
$ s_1 \approx s_2 $, 
\begin{enumerate}
	\item If $ \judgeDBreduce{s_1}{s_3} $,
	then
	there exist $ s_5 \in \mathcal{D} $, $ s_4 \in \mathcal{S}(C) $,
	such that \begin{itemize}
		\item $ s_5 \approx s_4 $, and
		\item $ \judgeSreduce{C}{s_2}{s_4} $,
		\item $ \judgeDBreduceTrans{s_3}{s_5} $
	\end{itemize}
	\item If $ \judgeSreduce{C}{s_2}{s_4} $,
	then
	there exist $ s_3 \in \mathcal{D} $,
	such that \begin{itemize}
		\item $ s_3 \approx s_4 $
		\item $ \judgeDBreduceTrans{s_1}{s_3} $
	\end{itemize}
\end{enumerate}
\end{lemma}
\begin{proof}
	See the supplementary material.
\end{proof}
